---
title: "Pubertal Timing & Tempo"
subtitle: "Individual Differences in the ABCD Cohort — Study 1"
author: "Clare McCann"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    echo: false
    warning: false
    message: false
    fig-align: center
    transition: slide
    footer: "McCann — Study 1"
    smaller: true
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

# export DATA_DIR="/u/project/silvers/data/ABCD/ABCD-release-6.0/cfm/physical-health/puberty"
# export OUT_DIR="/u/home/c/clarefmc/projects/abcd-projs/dissertation/study1/outputs"
# quarto render 05_slides.qmd

# Output path detection.
# On HPC: OUT_DIR env var points to the absolute outputs directory.
# Locally: use a path relative to this QMD file (scripts/ → ../outputs/).
out_base <- Sys.getenv("OUT_DIR")
if (!nzchar(out_base)) {
  out_base <- file.path("..", "outputs")
}
gamm_dir  <- file.path(out_base, "gamm")
lpa_dir   <- file.path(out_base, "lpa")
psych_dir <- file.path(out_base, "psychometrics")
reg_dir   <- file.path(out_base, "regression")
ds_names <- c("female_parent", "female_youth", "male_parent", "male_youth")

safe_read <- function(path) {
  if (!file.exists(path) || file.size(path) == 0) return(NULL)
  tryCatch(read.csv(path), error = function(e) NULL)
}

# Load timing/tempo summaries for sample size tables
tt_list <- lapply(ds_names, function(ds) {
  safe_read(file.path(gamm_dir, paste0(ds, "_timing_tempo.csv")))
})
names(tt_list) <- ds_names

# Load LPA best-solutions summary if available
lpa_summary_path <- file.path(lpa_dir, "lpa_best_solutions_summary.csv")
lpa_summary <- safe_read(lpa_summary_path)

# Helper: display image or placeholder text
img <- function(path, width = "90%") {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("*Output not yet available:* `", basename(path), "`"))
  }
}
```

# Background {background-color="#f5f5f5"}

## Puberty as a Process

- Puberty is not a single event — it unfolds across years with meaningful individual variation
- Two key dimensions:
  - **Timing** — when pubertal development begins
  - **Tempo** — how rapidly it progresses
- These are theoretically and empirically distinct yet often conflated

::: {.notes}
Motivate why we need to separate timing from tempo. Most prior work uses cross-sectional age-adjusted scores or single-timepoint assessments that can't disentangle these.
:::

## Study Aims

1. **Characterize population-level pubertal trajectories** using factor-smooth GAMMs
   - Fit a shared population curve + person-specific deviation in a single model (`mgcv::bam()`)
   - Separates *when* puberty occurs from *how fast* it progresses — without a two-stage approach

2. **Extract individual-level timing & tempo** estimates for each participant
   - **Timing deviation**: years earlier/later than cohort mean onset (PDS = 2.5 threshold)
   - **Acceleration at onset**: d²(PDS)/d(age²) at the onset point — how rapidly the growth rate is changing when puberty begins
   - **Peak velocity**: max d(PDS)/d(age) — retained for reference

3. **Identify latent pubertal profiles** via Latent Profile Analysis
   - Cluster on timing deviation and acceleration at onset (z-scored within sample)
   - `tidyLPA`/mclust: K = 1–6 profiles × 3 covariance structures (EEI, VVI, EEE)
   - Model selection by BIC (lower = better) + entropy (class separation ≥ 0.80)

4. [*Downstream: link profiles to outcomes — next script*]

::: {.notes}
Fill in aim 4 once downstream analyses are complete.
:::

# Sample {background-color="#f5f5f5"}

## ABCD Study — Pubertal Data

```{r sample-table}
sample_tbl <- bind_rows(lapply(names(tt_list), function(ds) {
  d <- tt_list[[ds]]
  if (is.null(d)) return(data.frame(Dataset = ds, N_participants = NA,
                                     N_with_timing = NA, Pct_timing = NA))
  data.frame(
    Dataset        = ds,
    N_participants = n_distinct(d$id),
    N_with_timing  = sum(!is.na(d$timing)),
    Pct_timing     = round(mean(!is.na(d$timing)) * 100, 1)
  )
}))

knitr::kable(
  sample_tbl,
  col.names = c("Dataset", "N participants", "N with timing estimate", "% with timing"),
  align     = "lrrr"
)
```

::: {.notes}
Participants without a timing estimate never crossed the onset threshold (PDS composite ≥ 2.0) within the observed age range — likely very early or very late developers relative to the measurement window.
:::

## Measures

- **PDS (Pubertal Development Scale)** — parent and youth report, annual waves
- **Composite score**: mean of 4 items (body hair, sex-specific item, skin changes, petb)
- Binary items (menarche / facial hair) modelled separately with binomial family
- Up to 7 waves (baseline through follow-up 6)

# Regression Characteristics {background-color="#f5f5f5"}

## What Is Item-Level Regression?

- PDS items are ordinal (1–4); puberty is biologically irreversible
- **Regression**: a participant's score on an item *decreases* from one wave to the next
- Apparent regression can arise from:
  - Measurement error / inattentive responding
  - **Caregiver switching** — a new reporter with no prior knowledge of the child
- Understanding regression rates informs how we treat the longitudinal composite

## Regression Prevalence

```{r reg-participant-table}
ps_path <- file.path(reg_dir, "reg_participant_summary.csv")
ps <- safe_read(ps_path)
if (!is.null(ps)) {
  ps %>%
    group_by(group) %>%
    summarise(
      N = n(),
      `% ever regressed` = round(mean(ever_regressed, na.rm = TRUE) * 100, 1),
      `Mean regression events` = round(mean(total_regressions, na.rm = TRUE), 2),
      `Mean waves with regression` = round(mean(n_waves_w_regression, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    knitr::kable(align = "lrrrr")
} else {
  cat("*Run 01_regression_characterization.R first.*")
}
```

## Item-Level Regression Rates

```{r item-reg-plot}
#| out-width: "95%"
img(file.path(reg_dir, "item_regression_rates.png"))
```

::: {.notes}
Each bar shows the % of wave-to-wave transitions where that item's score decreased. Higher rates suggest noisier items or items more susceptible to caregiver switching.
:::

## Regression Rate by Wave

```{r wave-reg-plot}
#| out-width: "95%"
img(file.path(reg_dir, "wave_regression_rates.png"))
```

::: {.notes}
Earlier waves may show higher rates if children are still in early puberty and scores fluctuate around the floor. Later waves may show higher rates due to increased caregiver switching.
:::

## Caregiver Switching

:::: {.columns}
::: {.column width="55%"}
```{r cg-reg-plot}
#| out-width: "100%"
img(file.path(reg_dir, "caregiver_switch_regression.png"))
```
:::
::: {.column width="45%"}
```{r cg-summary-table}
cg_path <- file.path(reg_dir, "caregiver_switch_summary.csv")
cg <- safe_read(cg_path)
if (!is.null(cg)) {
  knitr::kable(
    cg %>%
      mutate(
        `% ever switched` = round(pct_ever_switched * 100, 1),
        `Mean switches` = round(mean_switches, 2)
      ) %>%
      select(sex, reporter, n, `% ever switched`, `Mean switches`),
    align = "llrrr"
  )
} else {
  cat("*Run 01_regression_characterization.R first.*")
}
```
:::
::::

::: {.notes}
Regression coinciding with a caregiver switch is expected — the new caregiver doesn't know the child's prior pubertal status. Regression without a switch is more indicative of measurement error.
:::

## Predictors of Regression

```{r reg-model-table}
mod_path <- file.path(reg_dir, "reg_predictors_models.csv")
reg_mod <- safe_read(mod_path)
if (!is.null(reg_mod)) {
  reg_mod %>%
    filter(effect == "fixed", term != "(Intercept)") %>%
    select(group, term, estimate, conf.low, conf.high, p.value) %>%
    mutate(across(c(estimate, conf.low, conf.high), ~ round(., 2)),
           p.value = round(p.value, 3)) %>%
    knitr::kable(
      col.names = c("Dataset", "Predictor", "OR", "95% CI low", "95% CI high", "p"),
      caption = "Mixed logistic regression: predictors of any item regression (OR scale)"
    )
} else {
  cat("*Run 01_regression_characterization.R first.*")
}
```

# Psychometrics {background-color="#f5f5f5"}

## Measurement Validation — Overview

- **4 datasets**: female × {parent, youth}, male × {parent, youth}; up to 7 annual waves
- **Items**: peta (pubic hair), petb (growth spurt), petc (skin), petd (body odor) — all ordinal 1–4
- Four analyses conducted prior to GAMM modelling:

| Analysis | Question |
|---|---|
| GRM (per wave) | Which items discriminate best? |
| 1-factor CFA (per wave) | Does a single puberty factor fit? |
| Cross-reporter bifactor CFA | Are parent and youth measuring the same construct? |
| Longitudinal measurement invariance | Are factor loadings/thresholds stable across waves? |

## Item Discrimination — GRM

- Graded Response Model fit separately for each sex × reporter × wave
- Discrimination parameter *a* > 1.0 indicates good item–trait alignment
- Items flagged if *a* < 0.7 at the majority of waves

```{r disc-heatmap}
#| out-width: "90%"
img(file.path(psych_dir, "discrimination_heatmap.png"))
```

## Unidimensionality — CFA Fit

```{r cfa-fits}
cfa_path <- file.path(psych_dir, "cfa_fit_per_wave.csv")
cfa_fits <- safe_read(cfa_path)
if (!is.null(cfa_fits)) {
  knitr::kable(
    cfa_fits %>%
      group_by(dataset) %>%
      summarise(
        across(c(cfi.scaled, tli.scaled, rmsea.scaled, srmr),
               ~ round(mean(., na.rm = TRUE), 3)),
        .groups = "drop"
      ),
    col.names = c("Dataset", "CFI", "TLI", "RMSEA", "SRMR"),
    caption = "Average CFA fit across waves (WLSMV; target: CFI > .95, RMSEA < .06)"
  )
} else {
  cat("*Run 02_psychometrics.R first.*")
}
```

::: {.notes}
One-factor CFA with WLSMV estimator. Fit is averaged across waves; wave-level table in supplementary materials.
:::

## Cross-Reporter Structure — Bifactor CFA

- Three models compared per sex:
  - **M1 (method-only)**: reporter-specific method factors, no general trait — tests whether parent and youth are measuring the same thing at all
  - **M2 (bifactor)**: general puberty factor + freely correlated reporter method factors — the theoretically preferred MTMM model
  - **M3 (1-factor)**: all items load on one pooled puberty factor, no reporter distinction

```{r bifactor-table}
bf_path <- file.path(psych_dir, "bifactor_model_comparison.csv")
bf <- safe_read(bf_path)
if (!is.null(bf)) {
  knitr::kable(
    bf %>% select(sex, model, cfi.scaled, tli.scaled, rmsea.scaled, srmr) %>%
      mutate(across(where(is.numeric), ~ round(., 3))),
    col.names = c("Sex", "Model", "CFI", "TLI", "RMSEA", "SRMR")
  )
} else {
  cat("*Run 02_psychometrics.R first.*")
}
```

::: {.notes}
M2 is the theoretically motivated model: it tests whether parent and youth items share a common puberty construct (general factor) over and above reporter-specific variance (method factors). M1 fit vs. M3 fit tells you whether reporter distinction matters at all. For females, M2 fit excellent across waves; for males, M1 consistently fit better than M3, suggesting reporter disagreement is more pronounced in male puberty measurement.
:::

## Longitudinal Measurement Invariance

- Does the PDS factor structure hold across 7 annual waves?
- Sequence: configural → metric → scalar → strict (WLSMV, θ parameterization)
- Criterion: ΔCFI < −.010 indicates non-invariance

```{r long-inv-table}
li_path <- file.path(psych_dir, "longitudinal_invariance.csv")
li <- safe_read(li_path)
if (!is.null(li)) {
  knitr::kable(
    li %>%
      select(group, model, cfi.scaled, rmsea.scaled, delta_cfi) %>%
      mutate(across(where(is.numeric), ~ round(., 3))),
    col.names = c("Dataset", "Model", "CFI", "RMSEA", "ΔCFI")
  )
} else {
  cat("*Run 02_psychometrics.R first.*")
}
```

::: {.notes}
Scalar invariance (equal loadings + thresholds across waves) is required for valid comparison of latent means over time — i.e., for the GAMM trajectories to be interpretable as change in a common construct.
:::

# Analytic Approach {background-color="#f5f5f5"}

## Factor-Smooth GAMMs

$$\text{PDS}_{it} = f(\text{age}_t) + f_i(\text{age}_t) + \varepsilon_{it}$$

- $f(\text{age})$: **population mean trajectory** (shared smooth)
- $f_i(\text{age})$: **person-specific deviation** (factor smooth, `bs="fs"`)
- First-derivative penalty (`m=1`) shrinks individual curves toward the population mean
- Fitted via `mgcv::bam()` with `discrete=TRUE` for scalability at ABCD $N$

::: {.notes}
The factor-smooth approach avoids the limitation of two-stage methods (extract timing then regress) by estimating individual trajectories directly within a single model.
:::

## What We Extract

From each person's fitted trajectory $\hat{y}_i(\text{age})$:

| Construct | Definition |
|-----------|-----------|
| **Timing** | Age at which $\hat{y}_i$ first crosses PDS = 2.5 |
| **Timing deviation** | Timing − cohort mean timing; negative = early, positive = late |
| **Acceleration at onset** | $\frac{d^2\hat{y}_i}{dt^2}$ at onset — how rapidly the growth rate is changing at the moment puberty begins; positive = still accelerating, negative = already decelerating |
| **Peak velocity** | $\max_t \frac{d\hat{y}_i}{dt}$ — steepest rate of change (reference only) |

::: {.notes}
Acceleration at onset is the second derivative of the individual trajectory evaluated when they first cross PDS = 2.5. Negative values are biologically meaningful — they occur when someone crosses the threshold after their peak velocity, meaning their growth rate was already slowing. This tends to happen in later developers whose S-curve peak occurs before the PDS = 2.5 threshold is reached.
:::

# Population Trajectories {background-color="#f5f5f5"}

## Females — Population Smooth

:::: {.columns}
::: {.column width="50%"}
**Parent report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_parent_population_smooth.png"))
```
:::
::: {.column width="50%"}
**Youth report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_youth_population_smooth.png"))
```
:::
::::

## Males — Population Smooth

:::: {.columns}
::: {.column width="50%"}
**Parent report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_parent_population_smooth.png"))
```
:::
::: {.column width="50%"}
**Youth report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_youth_population_smooth.png"))
```
:::
::::

## Pubertal Velocity — Females

:::: {.columns}
::: {.column width="50%"}
**Parent report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_parent_growth_velocity.png"))
```
:::
::: {.column width="50%"}
**Youth report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_youth_growth_velocity.png"))
```
:::
::::

## Pubertal Velocity — Males

:::: {.columns}
::: {.column width="50%"}
**Parent report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_parent_growth_velocity.png"))
```
:::
::: {.column width="50%"}
**Youth report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_youth_growth_velocity.png"))
```
:::
::::

## Individual Trajectories — Females

:::: {.columns}
::: {.column width="50%"}
**Parent report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_parent_individual_trajectories.png"))
```
:::
::: {.column width="50%"}
**Youth report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_youth_individual_trajectories.png"))
```
:::
::::

::: {.notes}
300 randomly sampled participants. Colour = timing tertile (red = early, blue = on-time, green = late, grey = no onset detected within observation window). Bold black = population mean. Dashed line = onset threshold (PDS = 2.5).
:::

## Individual Trajectories — Males

:::: {.columns}
::: {.column width="50%"}
**Parent report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_parent_individual_trajectories.png"))
```
:::
::: {.column width="50%"}
**Youth report**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_youth_individual_trajectories.png"))
```
:::
::::

# Individual Timing & Tempo {background-color="#f5f5f5"}

## Distribution of Onset Age

:::: {.columns}
::: {.column width="50%"}
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_parent_timing_dist.png"))
```
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_parent_timing_dist.png"))
```
:::
::: {.column width="50%"}
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_youth_timing_dist.png"))
```
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_youth_timing_dist.png"))
```
:::
::::

## Distribution of Peak Velocity

:::: {.columns}
::: {.column width="50%"}
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_parent_peak_velocity_dist.png"))
```
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_parent_peak_velocity_dist.png"))
```
:::
::: {.column width="50%"}
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_youth_peak_velocity_dist.png"))
```
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_youth_peak_velocity_dist.png"))
```
:::
::::

## Timing vs. Peak Velocity

:::: {.columns}
::: {.column width="50%"}
**Females**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "female_parent_timing_vs_velocity.png"))
```
:::
::: {.column width="50%"}
**Males**
```{r}
#| out-width: "100%"
img(file.path(gamm_dir, "male_parent_timing_vs_velocity.png"))
```
:::
::::

::: {.notes}
If timing and peak velocity are moderately correlated, that motivates the LPA — we want to find people who are not just early/late but also fast/slow in ways that cluster naturally.
:::

# Latent Profile Analysis {background-color="#f5f5f5"}

## LPA Overview

- **Variables**: timing deviation (yr from cohort mean) + acceleration at onset (d²PDS/dt²), z-scored within sample
- **Method**: `tidyLPA` (wrapping `mclust`)
- **Covariance structures compared**: EEI (equal var, zero cov), VVI (varying var, zero cov), EEE (equal var + cov) — VVV excluded as overparameterized with 2 variables
- **Profiles searched**: K = 1–6
- **Selection criteria**: BIC (lower = better, primary) + entropy (> 0.80 = good separation)
- **Six solutions**: female × {parent, youth, averaged}, male × {parent, youth, averaged}

## Model Selection — Females

:::: {.columns}
::: {.column width="50%"}
**BIC**
```{r}
#| out-width: "100%"
img(file.path(lpa_dir, "female_parent_bic_plot.png"))
```
:::
::: {.column width="50%"}
**Entropy**
```{r}
#| out-width: "100%"
img(file.path(lpa_dir, "female_parent_entropy_plot.png"))
```
:::
::::

## Model Selection — Males

:::: {.columns}
::: {.column width="50%"}
**BIC**
```{r}
#| out-width: "100%"
img(file.path(lpa_dir, "male_parent_bic_plot.png"))
```
:::
::: {.column width="50%"}
**Entropy**
```{r}
#| out-width: "100%"
img(file.path(lpa_dir, "male_parent_entropy_plot.png"))
```
:::
::::

## Selected Solutions

```{r lpa-solution-table}
if (!is.null(lpa_summary)) {
  knitr::kable(
    lpa_summary,
    col.names = c("Sample", "K (profiles)", "Model"),
    align     = "lcc"
  )
} else {
  cat("*LPA outputs not yet available — run 04\\_lpa.R first.*")
}
```

## Pubertal Profiles — Females (Parent)

```{r}
#| out-width: "80%"
if (!is.null(lpa_summary)) {
  k <- lpa_summary$best_K[lpa_summary$label == "female_parent"]
  if (length(k) > 0)
    img(file.path(lpa_dir, paste0("female_parent_k", k, "_profile_plot.png")))
} else {
  cat("*Run 04\\_lpa.R to generate profile plots.*")
}
```

## Pubertal Profiles — Females (Averaged)

```{r}
#| out-width: "80%"
if (!is.null(lpa_summary)) {
  k <- lpa_summary$best_K[lpa_summary$label == "female_averaged"]
  if (length(k) > 0)
    img(file.path(lpa_dir, paste0("female_averaged_k", k, "_profile_plot.png")))
} else {
  cat("*Run 04\\_lpa.R to generate profile plots.*")
}
```

## Pubertal Profiles — Males (Parent)

```{r}
#| out-width: "80%"
if (!is.null(lpa_summary)) {
  k <- lpa_summary$best_K[lpa_summary$label == "male_parent"]
  if (length(k) > 0)
    img(file.path(lpa_dir, paste0("male_parent_k", k, "_profile_plot.png")))
} else {
  cat("*Run 04\\_lpa.R to generate profile plots.*")
}
```

## Pubertal Profiles — Males (Averaged)

```{r}
#| out-width: "80%"
if (!is.null(lpa_summary)) {
  k <- lpa_summary$best_K[lpa_summary$label == "male_averaged"]
  if (length(k) > 0)
    img(file.path(lpa_dir, paste0("male_averaged_k", k, "_profile_plot.png")))
} else {
  cat("*Run 04\\_lpa.R to generate profile plots.*")
}
```

# Summary {background-color="#f5f5f5"}

## Key Takeaways

::: {.incremental}
- Factor-smooth GAMMs recover individual pubertal trajectories from longitudinal PDS data
- Timing deviation and acceleration at onset are **negatively correlated** across profiles: earlier developers are still accelerating when puberty begins; later developers have already passed peak velocity
- LPA identifies **K = 6** distinct pubertal profiles (BIC-optimal, Model 2/VVI, entropy .87–.94) that differ on both *when* and *how fast* puberty is changing at onset — replicated across all six sex × reporter solutions
- Cross-reporter structure: both parent and youth load on a common puberty factor (M2 bifactor); reporter-specific variance present but structure is longitudinally stable (scalar ΔCFI < .001 for females; −.009 for males)
- Next: link profile membership to outcomes
:::

::: {.notes}
Fill in the bracketed text once LPA results are in and you've looked at the comparison plots.
:::

## Thank You

Questions?

\vspace{1em}

> *"Puberty is not just about when — it's about how fast."*

::: {.notes}
Optional closing quote placeholder — swap for something more fitting or remove.
:::
